---
title: "Online Regime Change Detection"
author: "Kyle Pelham, kyleplhm@gmail.com"
date: "2023-03-29"
output: 
  html_document: 
    theme: cerulean
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r}
library(tidyverse)
library(here)
library(ggplot2)
library(onlineBcp)
library(changepoint)
library(lubridate)
library(plotly)
```

# Introduction

This analysis is my contribution to a larger MSBA captstone project in support of a large North American Utilities company. Our group was tasked with detecting significant changes in constraint flow beyond normal seasonality also known as regime changes. The model used here is one of multiple used to detect these changes.

#### What is a constraint? What is constraint flow? Why is this important?

In the context of a utilities company, constraints refer to limitations in the power grid, impacting its efficiency and reliability. These constraints can arise from transmission line capacity, generation limitations, network topology, and regulatory or market restrictions. When combined with the term "constraint flow," it refers to the flow of electricity through a constrained part of the power grid, which can be influenced by these factors. Power companies and grid operators work to mitigate these constraints, ensuring stable and cost-effective electricity delivery.

Having a method to detect large changes in constraint flow can allow the utilities company to take proactive actions to maintain stability within the power grid
